# Variability of gene expression in SNs within defined subsets of genes

## Introduction

We previously (expression.variance.Rmd) determined that IPSDSNs show higher variability across samples in global gene expression than most other tissues. We would like to know whether this variability is restricted to certain subsets of genes.
I have also shown that certain GO categories related to neuronal function have higher variability than the rest of the genes.
Next, we're going to see whether genes that are "turned on" in IPSDSNs relative to IPSCs also have higher variability.

```{r Setup}
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
options(stringsAsFactors=F)

# This file is generated by compare.SN.GTEx.MDS.Rmd
outputDir <- "../results/gene_set_variance/"
rpkmThreshold <- 1
print(paste0("RPKM threshold for expressed genes: ", rpkmThreshold))
goSummariesFontSize = 12
```

```{r LoadFiles}
readRPKMTable = function(fname) {
  # The first 2 columns are gene_id and length
  df = readr::read_tsv(fname) %>% as.data.frame()
  # Fix gene names to remove the version number (e.g. ".4" in ENSG00000223972.4)
  df$gene_id <- gsub("\\.[\\d]+", "", df$gene_id, perl=T)
  rownames(df) <- df$gene_id
  # include gene_id
  df[, c(1, 3:ncol(df))]
}

sn.df.all = readRPKMTable("../../sn_hg37/sn.hg37.gencode_full.rpkm.txt.gz")
colnames(sn.df.all) = tolower(colnames(sn.df.all))

sn.meta.all = read.csv("../data/Metadata_All_Lines.csv", header=T)
sn.meta.all$Sample = tolower(sn.meta.all$Sample)
sn.meta = sn.meta.all %>% filter(! Sample %in% c("posc_1", "iakz_1", "koun_2", "yuze_1_1", "yuze_1_2"))
sn.df = sn.df.all %>% dplyr::select(-one_of(c("posc_1", "iakz_1", "koun_2", "yuze_1_1", "yuze_1_2")))

sn.meta.P2 <- sn.meta %>% filter(Protocol == "P2V2") %>% mutate(sampleDifferentiation=paste(Cell.line, Differentiaton.Replicate))
# Now remove extraction replicates, as these aren't interesting for
# investigating expression variability
sn.meta.P2 <- sn.meta.P2[!duplicated(sn.meta.P2$sampleDifferentiation),]

hipsci.rpkm.df <- readRPKMTable("../../hipsci/hipsci.hg37.gencode_full.rpkm.txt.gz")
hipsci.rpkm.df = hipsci.rpkm.df[, -1]
hipsci.rpkmavg.df = data.frame(gene_id = rownames(hipsci.rpkm.df), hipsci_mean_rpkm=rowMeans(hipsci.rpkm.df))
rm(hipsci.rpkm.df)
```


```{r CalculateRSDs}
mean_na_rm = function(x) { mean(x, na.rm=T) }

rownames(sn.df) = sn.df$gene_id
rpkm.mat.sn = as.matrix(sn.df[,sn.meta.P2$Sample])
rownames(rpkm.mat.sn) = sn.df$gene_id

rpkm.summary.sn = data.frame(gene_id = sn.df$gene_id, avgrpkm = apply(rpkm.mat.sn, 1, FUN=mean_na_rm))
rpkm.summary.sn$logavgrpkm = log(rpkm.summary.sn$avgrpkm)
ggplot(rpkm.summary.sn, aes(logavgrpkm)) +
  geom_histogram(binwidth = 0.2) +
  theme_gray(fontSize) +
  ggtitle("Sensory neuron gene RPKM")
numExpressed = sum(rpkm.summary.sn$avgrpkm > rpkmThreshold)
print(paste("Number of genes with mean RPKM across samples > ", rpkmThreshold, ":", numExpressed))
rpkm.summary.sn$cell_type = "IPSDSN"
rpkm.summary.sn$rsd = apply(rpkm.mat.sn, 1, FUN=function(x) {sd(x, na.rm=T) / mean_na_rm(x)})
rpkm.summary.sn = rpkm.summary.sn[order(-rpkm.summary.sn$rsd), ]
rpkm.summary.sn.x = rpkm.summary.sn[rpkm.summary.sn$avgrpkm > rpkmThreshold,]
rpkm.summary.sn.x = rpkm.summary.sn.x %>% inner_join(hipsci.rpkmavg.df, by="gene_id")
```


```{r Compare.DE.genes}
genesUpInSNs = read.delim("../results/ipsc_sn/deseq.iPSC.vs.SN.ff.turnedon.5x.txt")

rpkm.upinSN.df = rpkm.summary.sn.x %>% inner_join(genesUpInSNs, by="gene_id")
rpkm.highly.upinSN.df = rpkm.summary.sn.x %>% inner_join(genesUpInSNs, by="gene_id") %>% filter(log2FoldChange > log2(100), hipsci_mean_rpkm < 0.1)

rpkm.not.upinSN.df = rpkm.summary.sn.x %>% filter(!gene_id %in% rpkm.upinSN.df$gene_id)

median(rpkm.not.upinSN.df$avgrpkm)
median(rpkm.summary.sn.x$avgrpkm)
median(rpkm.upinSN.df$avgrpkm)
median(rpkm.highly.upinSN.df$avgrpkm)

summary(rpkm.not.upinSN.df$rsd)
summary(rpkm.summary.sn.x$rsd)
summary(rpkm.upinSN.df$rsd)
summary(rpkm.highly.upinSN.df$rsd)

plot.df = rpkm.summary.sn.x %>% dplyr::select(gene_id, rsd)
plot.df$geneset = "All genes"
plot.df = rbind(plot.df,
                rpkm.upinSN.df %>% dplyr::select(gene_id, rsd) %>% mutate(geneset = "Upregulated genes"))
#                rpkm.highly.upinSN.df %>% 
#                    dplyr::select(gene_id, rsd) %>% mutate(geneset = "Highly upregulated genes"))
plot.df$geneset = factor(plot.df$geneset, levels=c("All genes", "Upregulated genes"))

ggplot(plot.df, aes(log(rsd), fill=geneset)) +
  geom_density(alpha=0.7) +
  theme_bw(fontSize) +
  ggtitle("Sensory neuron gene CV") +
  xlab("log(CV)")


plot.df = rpkm.not.upinSN.df %>% dplyr::select(gene_id, rsd, logavgrpkm)
plot.df$geneset = "Remaining genes"
plot.df = rbind(plot.df,
                rpkm.upinSN.df %>% dplyr::select(gene_id, rsd, logavgrpkm) %>% mutate(geneset = "Upregulated 5x"))
plot.df$geneset = factor(plot.df$geneset, levels=c("Remaining genes", "Upregulated 5x"))

# Export at 6x4 for figure
ggplot(plot.df, aes(log10(rsd), fill=geneset)) +
  geom_density(alpha=0.7) +
  theme_bw(14) +
  xlab(expression(log[10](CV))) +
  scale_fill_manual(name="Gene set", values=c("#F68F00", "#56A4F9")) +
  theme(legend.position=c(0.9,0.9), legend.justification=c(0.8,0.8))
```

Clearly the gene which are upregulated (turned on) during differentiation are more highly variable.
Looking at the average expression levels of upregulated and remaining genes, this doesn't seem to be explained by expression level. But let's check that.

We can do a scatter plot of gene CV against expression level (RPKM). Here we see that upregulated genes are slightly more variable than remaining genes, even at the same expression levels.

```{r Compare.DE.genes.2}
ggplot(plot.df, aes(x=logavgrpkm, y=log10(rsd), col=geneset)) +
  geom_point(alpha=0.3) + geom_smooth() + theme_bw(14)
```

